Normalization by Evaluation for Sized Dependent Types

### 2017-06-08 nbe/icfp17.tex
[ ICFP 17 ] Factored out proof part into proof of Lemma 5.1(3)

\begin{lemma}[Closure properties for neutrals]
\label{lem:rclosne}
\bla
\begin{enumerate}
\item If\/ $\RNE \Gamma t n \piT U T$ and $\RD \Gamma u d U$ then $\RNE \Gamma {t\,u} {n\,d} {T[u]}$.
\item If\/ $\RNE \Gamma t n \piT \Size T$ and $\RSIZE \Gamma a \alpha$ then $\RNE \Gamma {t\,a} {n\,\alpha} {T[a]}$.
\item If\/ $\RNE \Gamma t n \forallT T$ and $\RSIZE {\resurrect\Gamma} a \alpha$ then $\RNE \Gamma {t\,\ann a} {n\,\ann \alpha} {T[a]}$.
\end{enumerate}
\end{lemma}

@@

\begin{enumerate}
\item
The premise is, after type conversion, $\Gamma \der t \doteq \tRne\, n : \piT U T$.
To demonstrate $\LR \Gamma t {T'} {\up {S'} n} {A'}$ we assume
$\xi : \Gamma' \leq \Gamma$ and $\LR{\Gamma'}{u}{U\xi} a A$ and show
$\LR{\Gamma'} {(t\xi\,u)} {T(\xi,u)} {\up{S[a]}(n\,\down A a)} {B[a]}$.
By induction hypothesis (\ref{it:out}) we have $\RD {\Gamma'} u {\down R a} {U\xi}$, and
together with the weakened premise $\Gamma' \der t\xi \doteq \tRne\, n : (\piT U T) \xi$ we get
$\Gamma' \der t\xi\, u \doteq \tRne\, (n\,\down R a) : T(\xi,u)$.
The goal follows by induction hypothesis (\ref{it:into}) for
% $B[a] \in s$.
$\LRTS{\Gamma'}{T(\xi,u)}{B[a]} s$.
% This follows by induction hypothesis (\ref{it:into}) for $B[a] \in \SET[\ell]$ if
% we can prove that for all $\xi' : \Gamma'' \leq \Gamma'$ there is a neutral normal form $m$
% such that $\RNe{\Gamma''} {(n\,\down A a)} m$ and $\Gamma'' \der (t\xi\,u)\xi' = m : T(\xi,u)\xi'$.
% Note that the last goal can be rewritten to $\Gamma'' \der t\xi\xi'\,(u\xi') = m : T(\xi\xi',u\xi')$.
% We know that $\RNe{\Gamma''} n {m_0}$ with $\Gamma'' \der t\xi\xi' = m_0 : T'\xi\xi'$.
% Thus, it is sufficient to show that there is some normal form $v$ such that
% $\R{\Gamma''} {\down A a} v$ and $\Gamma'' \der u\xi' = v : U\xi\xi'$.
% However, this follows by induction hypothesis (\ref{it:out}) on
% assumption $\LR{\Gamma'}{u}{U\xi} a A$ with $\xi'$.
\qed
\end{enumerate}

@@

\begin{enumerate}
\item We assume $\xi : \Gamma' \leq \Gamma$ and
$\resurrect{\Gamma'} \der a, b : \Size$ and $\alpha,\beta \in \SIZE$ with
$\LRSize {\resurrect{\Gamma'}} b \beta$ and show
$\LR {\Gamma'} {t\xi\ann a} {T(\xi,b)} {(\up {S'} n) \ann \alpha} {A[\beta]}$.
By the evaluation rules for reflection, it is sufficient to show
% It is sufficient to show
$\LR {\Gamma'} {t\xi\ann a} {T(\xi,b)} {\up {S[\alpha]} (n \ann \alpha)} {A[\beta]}$.
This follows by induction hypothesis (\ref{it:into}) if
$\LRTS {\Gamma'} {T(\xi,b)} {A[\beta]} s$ and $A[\beta] \shape S[\alpha]$ and
$\RNE {\Gamma'} {t\xi\ann a} {n \ann \alpha} {T(\xi,b)}$.
The first two of these subgoals are immediate.  For the third, assume
$\xi' : \Gamma'' \leq \Gamma'$.
We have to show that there exists a neutral normal form $m'$ such that
$\RNe {\Gamma''} {n \ann \alpha} m'$ and
$\Gamma'' \der (t\xi \ann a)\xi' = m' : (T(\xi,b))\xi'$.
Sizes can always be read back, thus, let $\RSize {\Gamma''} \alpha {a_0}$.
By assmption $\RNE \Gamma t n {T'}$, there is an $m$ with
$\RNe {\Gamma''} n m$ and
$\Gamma'' \der t\xi\xi' = m : T'\xi\xi'$.
The latter equation implies
$\Gamma'' \der t\xi\xi' \ann {a\xi'} = m \ann {a_0} : T(\xi\xi',b\xi')$
by irrelevant size application to
$\resurrect{\Gamma''} \der a\xi', a_0, b \xi' : \Size$,
thus we are done by choosing $m' : = m \ann{a_0}$.

\end{enumerate}

\end{caselist}
\end{proof*}
} % END LONGVERSION

>>>>

\begin{lemma}[Closure properties for neutrals]
\label{lem:rclosne}
\bla
\begin{enumerate}
\item If\/ $\RNE \Gamma t n \piT U T$ and $\RD \Gamma u d U$ then $\RNE \Gamma {t\,u} {n\,d} {T[u]}$.
\item If\/ $\RNE \Gamma t n \piT \Size T$ and $\RSIZE \Gamma a \alpha$ then $\RNE \Gamma {t\,a} {n\,\alpha} {T[a]}$.
\item \label{it:closneall}
If\/ $\RNE \Gamma t n \forallT T$ and $\resurrect\Gamma \der a,b : \Size$ and $\alpha \in \SIZE$ then $\RNE \Gamma {t\,\ann a} {n\,\ann \alpha} {T[b]}$.
%% TOO WEAK:
% \item If\/ $\RNE \Gamma t n \forallT T$ and $\RSIZE {\resurrect\Gamma} a \alpha$ then $\RNE \Gamma {t\,\ann a} {n\,\ann \alpha} {T[a]}$.
\end{enumerate}
\end{lemma}
\LONGVERSION{
\begin{proof*}
For (\ref{it:closneall}), assume
$\xi : \Gamma' \leq \Gamma$.
We have to show that there exists a neutral normal form $m'$ such that
$\RNe {\Gamma'} {n\, \ann \alpha} m'$ and
$\Gamma' \der (t\, \ann a)\xi = m' : T[b]\xi$.
Sizes can always be read back, thus, let $\RSize {\Gamma'} \alpha {a_0}$.
By assumption $\RNE \Gamma t n {\forallT T}$, there is an $m$ with
$\RNe {\Gamma'} n m$ and
$\Gamma' \der t\xi = m : (\forallT T)\xi$.
The latter equation implies
$\Gamma' \der t\xi \ann {a\xi} = m \ann {a_0} : T(\xi,b\xi)$
by irrelevant size application to
$\resurrect{\Gamma'} \der a\xi, a_0, b \xi : \Size$,
thus we are done with $m' := m \ann{a_0}$.
\qed
\end{proof*}
} % END LONGVERSION

@@

\begin{enumerate}
\item
The premise is, after type conversion, $\Gamma \der t \doteq \tRne\, n : \piT U T$.
To demonstrate $\LR \Gamma t {T'} {\up {S'} n} {A'}$ we assume
$\xi : \Gamma' \leq \Gamma$ and $\LR{\Gamma'}{u}{U\xi} a A$ and show
$\LR{\Gamma'} {(t\xi\,u)} {T(\xi,u)} {\up{S[a]}(n\,\down A a)} {B[a]}$.
By induction hypothesis (\ref{it:out}) we have $\RD {\Gamma'} u {\down R a} {U\xi}$, and
together with the weakened premise $\Gamma' \der t\xi \doteq \tRne\, n : (\piT U T) \xi$ we get
$\Gamma' \der t\xi\, u \doteq \tRne\, (n\,\down R a) : T(\xi,u)$.
The goal follows by induction hypothesis (\ref{it:into}) for
% $B[a] \in s$.
$\LRTS{\Gamma'}{T(\xi,u)}{B[a]} s$.
% This follows by induction hypothesis (\ref{it:into}) for $B[a] \in \SET[\ell]$ if
% we can prove that for all $\xi' : \Gamma'' \leq \Gamma'$ there is a neutral normal form $m$
% such that $\RNe{\Gamma''} {(n\,\down A a)} m$ and $\Gamma'' \der (t\xi\,u)\xi' = m : T(\xi,u)\xi'$.
% Note that the last goal can be rewritten to $\Gamma'' \der t\xi\xi'\,(u\xi') = m : T(\xi\xi',u\xi')$.
% We know that $\RNe{\Gamma''} n {m_0}$ with $\Gamma'' \der t\xi\xi' = m_0 : T'\xi\xi'$.
% Thus, it is sufficient to show that there is some normal form $v$ such that
% $\R{\Gamma''} {\down A a} v$ and $\Gamma'' \der u\xi' = v : U\xi\xi'$.
% However, this follows by induction hypothesis (\ref{it:out}) on
% assumption $\LR{\Gamma'}{u}{U\xi} a A$ with $\xi'$.
\end{enumerate}

@@

\begin{enumerate}
\item We assume $\xi : \Gamma' \leq \Gamma$ and
$\resurrect{\Gamma'} \der a, b : \Size$ and $\alpha,\beta \in \SIZE$ with
$\LRSize {\resurrect{\Gamma'}} b \beta$ and show
$\LR {\Gamma'} {t\xi\ann a} {T(\xi,b)} {(\up {S'} n) \ann \alpha} {A[\beta]}$.
By the evaluation rules for reflection, it is sufficient to show
% It is sufficient to show
$\LR {\Gamma'} {t\xi\ann a} {T(\xi,b)} {\up {S[\alpha]} (n \ann \alpha)} {A[\beta]}$.
This follows by induction hypothesis (\ref{it:into}) if
$\LRTS {\Gamma'} {T(\xi,b)} {A[\beta]} s$ and $A[\beta] \shape S[\alpha]$ and
$\RNE {\Gamma'} {t\xi\ann a} {n \ann \alpha} {T(\xi,b)}$.
The first two of these subgoals are immediate.
The third follows by Lemma~\ref{lem:rclosne} (\ref{it:closneall}) from the
weakened assumption
 $\RNE {\Gamma'} {t\xi} n {T'\xi}$.
% For the third, assume
% $\xi' : \Gamma'' \leq \Gamma'$.
% We have to show that there exists a neutral normal form $m'$ such that
% $\RNe {\Gamma''} {n \ann \alpha} m'$ and
% $\Gamma'' \der (t\xi \ann a)\xi' = m' : (T(\xi,b))\xi'$.
% Sizes can always be read back, thus, let $\RSize {\Gamma''} \alpha {a_0}$.
% By assmption $\RNE \Gamma t n {T'}$, there is an $m$ with
% $\RNe {\Gamma''} n m$ and
% $\Gamma'' \der t\xi\xi' = m : T'\xi\xi'$.
% The latter equation implies
% $\Gamma'' \der t\xi\xi' \ann {a\xi'} = m \ann {a_0} : T(\xi\xi',b\xi')$
% by irrelevant size application to
% $\resurrect{\Gamma''} \der a\xi', a_0, b \xi' : \Size$,
% thus we are done by choosing $m' : = m \ann{a_0}$.
%
\qed
\end{enumerate}
%
\end{caselist}
\end{proof*}
} % END LONGVERSION
